Nutritional Deficits and the Quantity-Quality Trade-off: Evidence from an Exogenous Fertility Shock in Low-Income Urban Settings in the Philippines

### 2025-07-06 docs/paper.Rmd
revise data and empirical strategy section
Updated the Data and Empirical Strategy section to clarify the primary use of DHS data, detail the empirical design, and specify the role of census and APIS data. The new text provides a more precise description of the identification strategy, sample construction, and robustness checks, and removes a prior citation. PDF and TeX files regenerated accordingly.

>>>>

\section{Data and Empirical Strategy}

The primary dataset used in this analysis is the Philippine Demographic and Health Surveys (DHS), specifically the 1993, 1998, 2003, 2008, and 2013 waves. These surveys are nationally representative and contain detailed, child-level anthropometric indicators, such as height-for-age z-scores (HAZ) and weight-for-height z-scores (WHZ). These metrics are internationally recognized as proxies for chronic and acute malnutrition, respectively. The DHS also provides rich information on maternal characteristics, household wealth, fertility histories, and geographic location, which enabled a quasi-experimental research design centered on the policy-induced fertility shock in Manila City.

The empirical strategy exploits the 2000 issuance of Executive Order No. 003, which effectively banned the distribution of modern contraceptives in public health centers across Manila City. This abrupt and localized withdrawal of contraceptive access resulted in a fertility increase concentrated within the city, but did not affect surrounding municipalities within the National Capital Region (NCR). Children born in Manila following the policy are considered exposed to a fertility shock, while children in other NCR cities over the same period serve as the control group. The analytical sample is restricted to children under five years old at the time of each survey round to ensure accurate measurement of early-life nutritional outcomes.

The identification approach employs a difference-in-differences (DiD) framework. The main specification is:

\[
\text{Nutrition}_{ict} = \alpha + \beta \cdot (\text{Post}_t \times \text{Manila}_c) + X_{ict}'\gamma + \lambda_c + \tau_t + \varepsilon_{ict},
\]

where \(\text{Nutrition}_{ict}\) denotes the HAZ or WHZ score for child \(i\) in city \(c\) born in year \(t\); \(\text{Post}_t\) is an indicator for children born after 2000; \(\text{Manila}_c\) is a treatment group dummy; and \(X_{ict}\) includes controls for child age in months, sex, birth order, maternal age and education, household wealth quintile, and urban barangay indicators. City and year fixed effects \((\lambda_c, \tau_t)\) are included to account for time-invariant differences and common macroeconomic shocks. Standard errors are clustered at the city-year level to address potential autocorrelation and heteroskedasticity.

The 1998 DHS wave is treated as the pre-policy benchmark, while 2003 serves as the primary post-policy round for estimating short-run effects. The 2008 survey is incorporated into robustness checks to evaluate medium-run persistence of any observed nutritional gaps. The 1993 DHS is used in placebo tests to evaluate pre-policy parallel trends between Manila and control cities. Although included cautiously due to possible confounding from attrition and mobility, the 2013 DHS is optionally used to examine longer-term cohort outcomes and nutritional trajectories.

To validate the existence and magnitude of the fertility shock itself, and to assess potential violations of the identifying assumptions, Philippine Census microdata from 1990, 1995, 2000, 2007, and 2010 are used. The 1995 wave is treated as the pre-policy reference period for fertility, household size, and demographic composition, while 2007 captures the post-policy landscape. The 2000 census wave is excluded from baseline comparisons due to uncertainty about when contraceptive access began declining, though it is retained for robustness checks involving changes in fertility behavior. The 2010 wave is used to examine longer-run demographic patterns. A placebo comparison between 1990 and 1995 confirms that fertility rates in Manila and comparison cities followed parallel trajectories prior to the policy change.

To supplement both the DHS and Census, the Annual Poverty Indicators Survey (APIS) is used to provide household-level context on food expenditure, maternal education, and living standards. While more limited in sample size and coverage, the APIS enables exploratory analyses of heterogeneity in treatment effects by wealth and education, and supports the triangulation of maternal resource constraints during the post-policy period.

### 2025-07-06 docs/paper.tex
revise data and empirical strategy section
Updated the Data and Empirical Strategy section to clarify the primary use of DHS data, detail the empirical design, and specify the role of census and APIS data. The new text provides a more precise description of the identification strategy, sample construction, and robustness checks, and removes a prior citation. PDF and TeX files regenerated accordingly.

>>>>

\section{Data and Empirical Strategy}

The primary dataset used in this analysis is the Philippine Demographic
and Health Surveys (DHS), specifically the 1993, 1998, 2003, 2008, and
2013 waves. These surveys are nationally representative and contain
detailed, child-level anthropometric indicators, such as height-for-age
z-scores (HAZ) and weight-for-height z-scores (WHZ). These metrics are
internationally recognized as proxies for chronic and acute
malnutrition, respectively. The DHS also provides rich information on
maternal characteristics, household wealth, fertility histories, and
geographic location, which enabled a quasi-experimental research design
centered on the policy-induced fertility shock in Manila City.

The empirical strategy exploits the 2000 issuance of Executive Order
No.~003, which effectively banned the distribution of modern
contraceptives in public health centers across Manila City. This abrupt
and localized withdrawal of contraceptive access resulted in a fertility
increase concentrated within the city, but did not affect surrounding
municipalities within the National Capital Region (NCR). Children born
in Manila following the policy are considered exposed to a fertility
shock, while children in other NCR cities over the same period serve as
the control group. The analytical sample is restricted to children under
five years old at the time of each survey round to ensure accurate
measurement of early-life nutritional outcomes.

The identification approach employs a difference-in-differences (DiD)
framework. The main specification is:

\[
\text{Nutrition}_{ict} = \alpha + \beta \cdot (\text{Post}_t \times \text{Manila}_c) + X_{ict}'\gamma + \lambda_c + \tau_t + \varepsilon_{ict},
\]

where \(\text{Nutrition}_{ict}\) denotes the HAZ or WHZ score for child
\(i\) in city \(c\) born in year \(t\); \(\text{Post}_t\) is an
indicator for children born after 2000; \(\text{Manila}_c\) is a
treatment group dummy; and \(X_{ict}\) includes controls for child age
in months, sex, birth order, maternal age and education, household
wealth quintile, and urban barangay indicators. City and year fixed
effects \((\lambda_c, \tau_t)\) are included to account for
time-invariant differences and common macroeconomic shocks. Standard
errors are clustered at the city-year level to address potential
autocorrelation and heteroskedasticity.

The 1998 DHS wave is treated as the pre-policy benchmark, while 2003
serves as the primary post-policy round for estimating short-run
effects. The 2008 survey is incorporated into robustness checks to
evaluate medium-run persistence of any observed nutritional gaps. The
1993 DHS is used in placebo tests to evaluate pre-policy parallel trends
between Manila and control cities. Although included cautiously due to
possible confounding from attrition and mobility, the 2013 DHS is
optionally used to examine longer-term cohort outcomes and nutritional
trajectories.

To validate the existence and magnitude of the fertility shock itself,
and to assess potential violations of the identifying assumptions,
Philippine Census microdata from 1990, 1995, 2000, 2007, and 2010 are
used. The 1995 wave is treated as the pre-policy reference period for
fertility, household size, and demographic composition, while 2007
captures the post-policy landscape. The 2000 census wave is excluded
from baseline comparisons due to uncertainty about when contraceptive
access began declining, though it is retained for robustness checks
involving changes in fertility behavior. The 2010 wave is used to
examine longer-run demographic patterns. A placebo comparison between
1990 and 1995 confirms that fertility rates in Manila and comparison
cities followed parallel trajectories prior to the policy change.

To supplement both the DHS and Census, the Annual Poverty Indicators
Survey (APIS) is used to provide household-level context on food
expenditure, maternal education, and living standards. While more
limited in sample size and coverage, the APIS enables exploratory
analyses of heterogeneity in treatment effects by wealth and education,
and supports the triangulation of maternal resource constraints during
the post-policy period.
